### 2019-03-17 README.Rmd
update readme
updating README
also includes work on vignettes, articles

The goal of `taxadb` is to provide fast access to taxonomic data and manipulations, such as resolving taxonomic names to ids, looking up higher classification ranks of given species, or returning a list of all species below a given rank. These tasks are particularly common when synthesizing data across large species assemblies, such as combining occurrence records with trait records.

@@

Before we can use most `taxadb` functions, we need to do a one-time installation of the database `taxadb` uses for almost all commands.  This can take a while to run, but needs only be done once.  The database is installed on your local hard-disk and will persist between R sessions.  By default, the database will be installed in your user's application data directory, as detected by the `rappdirs` package.  (Set a custom location instead using `dbdir` argument.)

```{r }
library(taxadb)
td_create()
```

The default behavior installs only the ITIS database.  You can also specify a list of authorities to install, or install every authority using `td_create("all")`.

## Test drive

Once the databases have been set up, we're ready to explore.  
Here's a list of all the birds species known to ITIS:

```{r}
descendants(name = "Aves", rank = "class")
```

## Learn More

### [An introduction to taxadb](https://cboettig.github.io/taxadb/articles/articles/taxadb.html)

The [taxadb introduction](https://cboettig.github.io/taxadb/articles/articles/taxadb.html) 
provides an overview showing how `taxadb` functions can help us synthesize data across a given list of species by resolving synonyms and identifiers.

### [taxadb schemas](https://cboettig.github.io/taxadb/articles/articles/schema.html)

See the [schemas](https://cboettig.github.io/taxadb/articles/articles/schema.html) vignette for an overview of the underlying tables used by `taxadb` functions, and more about the different authorities accessed by taxadb.

```{r include=FALSE}
taxadb:::td_disconnect()
MonetDBLite::monetdblite_shutdown()
```

>>>>

The goal of `taxadb` is to provide *fast*, *consistent* access to taxonomic data, supporting commont tasks such as resolving taxonomic names to identifiers, looking up higher classification ranks of given species, or returning a list of all species below a given rank. These tasks are particularly common when synthesizing data across large species assemblies, such as combining occurrence records with trait records.

@@

```{r}
library(taxadb)
library(dplyr)
library(readr)
```

Create a local copy of the Catalogue of Life (2018) database:

```{r }
td_create("col")
```

Read in the species list used by the Breeding Bird Survey:

```{r, message = FALSE}
bbs_species_list <- system.file("extdata/bbs.tsv", package="taxadb")
bbs <- read_tsv(bbs_species_list)
```

## Getting names and ids

Two core functions are `get_ids()` and `get_names()`.  These functions take a vector of names or ids (respectively), and return a vector of ids or names (respectively).  For instance, we can use this to attempt to resolve all the bird names in the Breeding Bird Survey against the Catalogue of Life:

```{r}
birds <- bbs %>% 
  select(species) %>% 
  mutate(id = get_ids(species, "col"))

head(birds, 10)
```

Note that some names cannot be resolved to an identifier.  This can occur because of mis-spellings, non-standard formatting, or the use of a synonym not recognized by the naming provider.  Names that cannot be uniquely resolved because they are known synonyms of multiple different species will also return `NA`.  The `by_name` filtering functions can help us resolve this last case (see below).

`get_ids()` returns the IDs of accepted names, that is `dwc:AcceptedNameUsageID`s.  We can resolve the IDs into accepted names:

```{r}
birds %>% 
  mutate(accepted_name = get_names(id, "col"))
```

This illustrates that some of our names, e.g. *Dendrocygna bicolor* are accepted in the Catalogue of Life, while others, *Anser canagicus* are **known synonyms** of a different accepted name: **Chen canagica**.  Resolving synonyms and accepted names to identifiers helps us avoid the possible miss-matches we could have when the same species is known by two different names.

## Taxonomic Data Tables

Local access to taxonomic data tables lets us do much more than look up names and ids.  A family of `by_*` functions in `taxadb` help us work directly with subsets of the taxonomic data.  As we noted above, this can be useful in resolving certain ambiguous names.

For instance, *Trochalopteron henrici gucenense* does not resolve to an identifier in ITIS:

```{r}
get_ids("Trochalopteron henrici gucenense") 
```

Using `by_name()`, we find this is because the name resolves not to zero matches, but to more than one match:

```{r}
by_name("Trochalopteron henrici gucenense") 
```

```{r}
by_name("Trochalopteron henrici gucenense")  %>%
  mutate(acceptedNameUsage = get_names(acceptedNameUsageID)) %>% 
  select(scientificName, taxonomicStatus, acceptedNameUsage, acceptedNameUsageID)
```

Similar functions `by_id`, `by_rank`, and `by_common` take IDs, scientific ranks, or common names, respectively.  Here, we can get taxonomic data on all bird names in the Catalogue of Life:

```{r}
by_rank(name = "Aves", rank = "class", provider = "col")
```

Combining these with `dplyr` functions can make it easy to explore this data: for instance, which families have the most species?

```{r}
by_rank(name = "Aves", rank = "class", provider = "col") %>%
  filter(taxonomicStatus == "accepted", taxonRank=="species") %>% 
  group_by(family) %>%
  count(sort = TRUE) %>% 
  head()
```

## Using the database connection directly

`by_*` functions by default return in-memory data frames.  Because they are filtering functions, they return a subset of the full data which matches a given query (names, ids, ranks, etc), so the returned data.frames are smaller than the full record of a naming provider.  Working directly with the SQL connection to the MonetDBLite database gives us access to all the data. The `taxa_tbl()` function provides this connection:

```{r}
taxa_tbl("col")
```

We can still use most familiar `dplyr` verbs to perform common tasks.  For instance: which species has the most known synonyms?

```{r}
most_synonyms <- taxa_tbl("col") %>% 
  group_by(acceptedNameUsageID) %>% 
  count(sort=TRUE)
most_synonyms
```
However, unlike the `by_*` functions which return convienent in-memory tables, this is still a remote connection.  This means that direct access using the `taxa_tbl()` function (or directly accessing the database connection using `td_connect()`) is more low-level and requires greater care.  For instance, we cannot just add a `%>% mutate(acceptedNameUsage = get_names(acceptedNameUsageID))` to the above, becuase `get_names` does not work on a remote collection.  Instead, we would first need to use a `collect()` to pull the summary table into memory.  Users familiar with remote databases in `dplyr` will find using `taxa_tbl()` directly to be convenient and fast, while other users may find the `by_*` approach to be more intuitive.

So which species had those 456 names?

```{r}
most_synonyms %>% 
  head(1) %>% 
  pull(acceptedNameUsageID) %>% 
  by_id("col") %>%
  select(scientificName)
```

```{r include=FALSE}
taxadb:::td_disconnect()
MonetDBLite::monetdblite_shutdown()

if(require(codemetar)) codemetar::write_codemeta()
```
